The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free eLearning program CrypTool\\ (9th edition -- distributed with CrypTool version 1.4.20)} \vspace{60pt}

### 2009-07-01 script/en/chapters/paper_and_pencil.tex
Add Sage examples for chapter on paper-pencil ciphers

@@

>>>>

\bibitem[Hill1929]{Hill1929} \index{Hill 1929}
   Lester S. Hill,\\
   ``Cryptography in an Algebraic Alphabet,''
   \emph{The American Mathematical Monthly}, 36(6):306--312, 1929.

\bibitem[Hill1931]{Hill1931} \index{Hill 1931}
   Lester S. Hill,\\
   ``Concerning Certain Linear Transformation Apparatus of Cryptography,''
   \emph{The American Mathematical Monthly}, 38(3):135--154, 1931.

@@

% ---------------------------------------------------------------------------
\newpage
\section*{Appendix A: Examples using Sage}
\addcontentsline{toc}{section}{Appendix A: Examples using Sage}
\index{Sage}

\subsection*{Transposition ciphers using Sage}

Tranposition ciphers are implemented in the class
\begin{center}
\verb!sage.crypto.classical.TranspositionCryptosystem!
\end{center}
To construct and work with a transposition cipher, we first need to
determine the alphabet that contains our plaintext and ciphertext
spaces. Typically, this alphabet will be the upper-case letters of the
English alphabet, which can be accessed via the function
\begin{center}
\verb!sage.monoids.string_monoid.AlphabeticStrings!
\end{center}
We then need to decide on the block length of a block permutation,
which is the length of the row vector to be used in encryption and
decryption. This row vector is our key, and it specifies a permutation
of a plaintext.

Here is a transposition cipher of block length 14, where every letter
in the plaintext is shifted to the left by one character, with wrap
around. That is the encryption process. The decryption process is
shifting to the left by $14 - 1 = 13$ characters in order to recover
the plaintext.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher of block length 14
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage: # our encryption/decryption key
sage: key = [2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1]
sage:
sage: # encrypt a plaintext
sage: E = T(key)
sage: C = E(T.encoding("a b c d e f g h i j k l m n")); C
BCDEFGHIJKLMNA
sage:
sage: # decrypt ciphertext
sage: keyInv = [14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13]
sage: D = T(keyInv)
sage: D(C)
ABCDEFGHIJKLMN
\end{Verbatim}

The encryption key can also be such that it shifts each letter of the
plaintext to the left by two characters, with wrap around. Then we
need to shift each letter of the ciphertext to the left by $14 - 2 =
12$ characters in order to obtain the plaintext.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage: # encrypt by shifting to the left by 2 letters
sage: key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage: E = T(key)
sage: C = E(T.encoding("a b c d e f g h i j k l m n")); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt by shifting to the left by 12 letters
sage: keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: D = T(keyInv)
sage: D(C)
ABCDEFGHIJKLMN
\end{Verbatim}

As a more complicated example, we can use a permutation in the
encryption and decryption processes. If the block length of our
transposition cipher is $n$, then the encryption/decryption
key must be a permutation on $n$ symbols. The following
example uses the method \verb!random_key()! of the class
\verb!TranspositionCryptosystem!. So each call to
\verb!random_key()! produces a different key for encryption, which
affects the ciphertext. Note that your results may be different from
the following example.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage:
sage: # encryption
sage: key = T.random_key(); key
(1,2,7)(3,9)(4,5,10,12,8,13,11)(6,14)
sage: P = T.encoding("a b c d e f g h i j k l m n")
sage: C = T.enciphering(key, P); C
BGIEJNAMCLDHKF
sage:
sage: # decryption
sage: T.deciphering(key, C)
ABCDEFGHIJKLMN
sage: T.deciphering(key, T.enciphering(key, P))
ABCDEFGHIJKLMN
\end{Verbatim}

% ---------------------------------------------------------------------------

\subsection*{Substitution ciphers}

Substitution cryptosystems are implemented in Sage in the class

\begin{center}
\verb!sage.crypto.classical.SubstitutionCryptosystem!
\end{center}

\noindent The following code sample uses Sage to construct a substitution
cipher with a random key. A random key can be generated using the
method \verb!random_key()! of the class
\verb!SubstitutionCryptosystem!. Different keys determine different
substitution ciphers. So each call to \verb!random_key()! returns
different results. First, we perform the encryption process with a
random key.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # the plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage:
sage: # substitution cipher with random key
sage: S = SubstitutionCryptosystem(A)
sage: key = S.random_key(); key
CMPGOWFIJRESXNYBKUDALQHTVZ
sage:
sage: # encryption
sage: msg = A.encoding("Substitute this with something else better."); msg
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: S.enciphering(key, msg)
DLMDAJALAOAIJDHJAIDYXOAIJNFOSDOMOAAOU
\end{Verbatim}

We now decrypt the ciphertext to recover our plaintext.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # decryption
sage: plaintext = S.deciphering(key, ciphertext); plaintext
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: plaintext == msg
True
\end{Verbatim}

% ---------------------------------------------------------------------------

\subsection*{Caesar cipher}

The following example uses Sage to construct a Caesar
cipher. First, we demonstrate encryption using this cipher.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
....: 20, 21, 22, 23, 24, 25, 0, 1, 2])
sage:
sage: # encrypt message
sage: encrypt = S(key)
sage: msg = A.encoding("Shift three positions to the right."); msg
SHIFTTHREEPOSITIONSTOTHERIGHT
sage: ciphertext = encrypt(msg); ciphertext
VKLIWWKUHHSRVLWLRQVWRWKHULJKW
\end{Verbatim}

Next, we recover the plaintext.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # decrypt message
sage: keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
....: 14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: decrypt = S(keyInv)
sage: plaintext = decrypt(ciphertext); plaintext
SHIFTTHREEPOSITIONSTOTHERIGHT
sage: plaintext == msg
True
\end{Verbatim}

% ---------------------------------------------------------------------------

\subsection*{Substitution with symbols}

Here is a Sage example where the symbols are from the binary number
system. There is very little security when using this system as
our plaintext/ciphertext alphabet because the binary number system
only has two elements: 0 and 1.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # the plaintext/ciphertext alphabet
sage: A = BinaryStrings()
sage: # substitution cipher over the alphabet A
sage: S = SubstitutionCryptosystem(A); S
Substitution cryptosystem on Free binary string monoid
sage: key = S.random_key(); key
10
sage:
sage: # encryption
sage: msg = A.encoding("Working with the binary numbers."); msg
01010111011011110111001001101011011010010110111001100111001000000111011101101\
00101110100011010000010000001110100011010000110010100100000011000100110100101\
10111001100001011100100111100100100000011011100111010101101101011000100110010\
1011100100111001100101110
sage: ciphertext = S.enciphering(key, msg); ciphertext
10101000100100001000110110010100100101101001000110011000110111111000100010010\
11010001011100101111101111110001011100101111001101011011111100111011001011010\
01000110011110100011011000011011011111100100011000101010010010100111011001101\
0100011011000110011010001
sage:
sage: # decryption
sage: plaintext = S.deciphering(key, ciphertext); plaintext
01010111011011110111001001101011011010010110111001100111001000000111011101101\
00101110100011010000010000001110100011010000110010100100000011000100110100101\
10111001100001011100100111100100100000011011100111010101101101011000100110010\
1011100100111001100101110
sage: plaintext == msg
True
\end{Verbatim}

However, our plaintext/ciphertext space  can be over a larger
alphabet, say the hexadecimal number system. Here is an example
using the hexadecimal number system as our substitution alphabet.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: A = HexadecimalStrings()
sage: S = SubstitutionCryptosystem(A)
sage: key = S.random_key(); key
26849de70f35b1ca
sage: msg = A.encoding("Working with a larger alphabet."); msg
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: ciphertext = S.enciphering(key, msg); ciphertext
d7ea78e5efece78277ef79e082e682ebe678e7ed7882e6eb72e0e6e8ed798c
sage: plaintext = S.deciphering(key, ciphertext); plaintext
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: plaintext == msg
True
\end{Verbatim}

% ---------------------------------------------------------------------------

\subsection*{Vigen{\`e}re cipher}

The Vigen{\`e}re cipher is implemented in the Sage class

\begin{center}
\verb!sage.crypto.classical.VigenereCryptosystem!
\end{center}

\noindent For our ciphertext/plaintext space, we can work with the upper-case
letters of the English alphabet, the binary number system, the
octal number system, or the hexadecimal number system. Here is an
example using the class \verb!AlphabeticStrings!, which implements
the English capital letters.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # construct Vigenere cipher
sage: A = AlphabeticStrings()
sage: V = VigenereCryptosystem(A, 19); V
Vigenere cryptosystem on Free alphabetic string monoid on A-Z of
period 19
sage: key = V.random_key(); key
ETKAAYJBKURHAUJXOMC
sage:
sage: # encryption
sage: msg = V.encoding("The Vigenere cipher is polyalphabetic."); msg
THEVIGENERECIPHERISPOLYALPHABETIC
sage: ciphertext = V.enciphering(key, msg); ciphertext
XAOVIENOOLVJIJQBFUUTHVYAJYIKVVAIW
sage:
sage: # decryption
sage: plaintext = V.deciphering(key, ciphertext); plaintext
THEVIGENERECIPHERISPOLYALPHABETIC
sage: plaintext == msg
True
\end{Verbatim}

% ---------------------------------------------------------------------------

\subsection*{Hill cipher}

The Hill~\cite{Hill1929,Hill1931} or matrix cipher is more
mathematically sophisticated than other ciphers mentioned in this
chapter. The encryption/decryption key of this cipher is an invertible
square matrix, and the plaintext/ciphertext is also a matrix. The
encryption and decryption processes use matrix operations. The Hill
cipher is implemented in the Sage class

\begin{center}
\verb!sage.crypto.classical.HillCryptosystem!
\end{center}

Here is an example where our plaintext/ciphertext space is the
capital letters of the English alphabet. In the Hill cipher, each
letter of this alphabet is assigned a unique integer modulo 26.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # construct a Hill cipher
sage: A = AlphabeticStrings()
sage: H = HillCryptosystem(A, 19); H
Hill cryptosystem on Free alphabetic string monoid on A-Z of block
length 19
sage: key = H.random_key(); key

[ 6 25  1 12  6 20 22 25  8 22 14 19 19  5  9 21 22  1 22]
[ 3  2  7 17 14 17  5 11 20  6  5  1 14 14 21 18  1 20  4]
[ 6 15 24  7  4 23 23 16 12  3 11 20 21  2  2 10 16  4 17]
[ 8 19  9 19 20  6 12  1 17 11 25 24 14 10  2 15  4  3 25]
[25 24 15 12  6  7 12 10  7 17 16 11 11 18 13 24 18 25 20]
[16 11 11 12  3  6 14 25 24  3  4 14 13 14 21  1  5  1 24]
[25  9  0 16  2 22  8  3  1  4 12 22 22 14  0  3 20  2 15]
[ 7 19 17 25 10 22  8  9  0  5 25  2 17 23  3 18 18  7  9]
[15  7  8 12  0 12 24 23 18 21  9 24 24 20 13 13  8 24  9]
[11 21 11 25  0  2 23 25  2  8  1  9 20  7 14  1 24  4 10]
[21  8 14  8 12 24 14  2 19  1  7 19  1 10 24  5  9  1  9]
[ 4  5 23 24  3  4 17 22  3 12  1 12  0 21 12  3 12 18  6]
[22 25 16 15 14  1  1 23  7 24  8 14 14 18 17 23 20 23 22]
[11  6 16  5  3  6 14  8 13 17 12 21  3 16  0  7  7 14 14]
[18 21  5 14 13 12  7  9  2 17 11 20 25 12  7 18 19 20  1]
[18 20 19  9  8  7 18 21 15 17 17 20 15 11  0  0  4 18 24]
[ 3  4 18  9 18  4  9 21 23 25  0  2 22 16  9  0 23 20 12]
[20 21 14  6  0  6 15 19 20 19 17  1 18 12  5 11  0  2 10]
[ 3  0 13 18 24 22 19  1 19 18  8 23  3 13 15 16  6 17  3]
sage:
sage: # encryption
sage: msg = H.encoding("Hill or matrix cipher uses matrix operations.")
sage: ciphertext = H.enciphering(key, msg); ciphertext
GJLILYPELTCHRKNXSTGJYCZSVNMPTIUZCTOJPU
sage:
sage: # decryption
sage: plaintext = H.deciphering(key, ciphertext); plaintext
HILLORMATRIXCIPHERUSESMATRIXOPERATIONS
sage: plaintext == msg
True
\end{Verbatim}
